Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-04-16 paper/paper.qmd
edit according to TA review

\newpage
# Introduction
Over the last few decades, the United States have implemented many policies, in an effort to address and mitigate poverty [@Kearney_Harris]. In 2019, the United States experienced the culmination of an economic expansion that followed the Great Recession. The robustness of the U.S. economy during this period was evidenced by low poverty rates, which was estimated to be 10.5% in 2019, down from 11.8% in 2018, marking the fifth consecutive year of decline [@Semega_2022]. In order to reduce poverty further, it is important for us to analyze why some US household are in poverty in 2019. We analyze the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of official poverty measure critics, providing an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. We primarily look at two factors: mortgage status and income in each household. This allows us to look at the question whether potential financial factors may cause the poverty status for some US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various financial factors captured in the SPM dataset. We select 17 data features for analysis. After some data cleaning, we summarize 17 features into 2 variables: 1) mortgage state and 2) household annual total income after tax, plus government benefits and minus expenses. The estimand in this paper is the number of households who are in poverty in reality. However, it is difficult to measure the exact number of households who are in poverty since there are millions of people in the United States and not all of them will be assessed due to various difficulties. For example, due to regional issues, some people who live in rural areas are not included in the survey. Therefore, in this paper, we attempt to estimate the estimand using a logistic regression model which is trained using sample dataset from the 2019 Revised Supplemental Poverty Measure Research Data from USCS.

@@

@@

## Features
The original SPM 2019 dataset, which shows in @tbl-raw, contains 157959 data entries and many variables. Since it is difficult to observe such a large dataset, this report will only explore and analyze through several data features. We first chose these 17 variables: h_seq, spm_poor, spm_tenmortstatus, spm_totval, spm_snapsub, spm_caphousesub, spm_schlunch, spm_engval, spm_wicval, spm_fedtax, spm_eitc, spm_actc, spm_fica, spm_sttax, spm_childsuppd, spm_capwkccxpns, and spm_medxpns. Their explanation are in @tbl-feature. This is because we want to calculate the income after tax, plus all the subsidies and benefits, minus all the expenses, just like how the SPM does it [@sgp2019]. If we only choose income before tax to analyze, then our analysis will be like the official poverty measure, which requires improvement.

@@

```{r}
#| label: tbl-raw
#| tbl-cap: "Preview of the raw 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false

@@

@@

First, a new column "income after tax" is created by summing up all incomes, credits, and benefits and subtracting all expenses. Then the dataset is cleaned by renaming of column headers, defining column classes, grouping the income variable into different income levels (here the assumed facts are that the income unit is in USD and 'k' stands for thousand, eg. 5k = 10,000), and replacing numerical values in the table with their corresponding descriptions from the data dictionary to improve the readability. After cleaning, 62990 rows of data with 3 data features remain. @tbl-clean-data shows a preview of the cleaned dataset.

@@

```{r}
#| label: tbl-clean-data
#| tbl-cap: "Preview of the cleaned 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false

@@

```{r}
#| label: tbl-summary
#| tbl-cap: "Statistics summary of the cleaned 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false

@@

```{r}
#| label: fig-mortgage
#| fig-cap: The distribution of poverty status by mortgage state
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = mortgage_state, fill = poverty_status)) +
  geom_bar(position = "dodge", width = 0.5) +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Mortgage state",
    y = "Number of households",
    fill = "Poverty Status"
  )
```

@@

```{r}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false

@@

## Model Implication
For posterior predictive checks, in @fig-post_dist, the great fit of the posterior distribution from our logistic regression model with actual poverty data suggests accurate capture of poverty status patterns. This indicates the accuracy of our model's poverty status prediction to the 2019 SPM data. Also, @fig-post_prior compares the posterior to the prior, which shows some parameter changes such as "below 5k" and "5k-50k", as well as the intercept (people who are owners with mortgage and earn 100-150k). The discrepancy for the parameter suggest that our prior may not be very accurate in regards to these specific aspects.

More detailed explanation of each plot can be found in Appendix @sec-model-details.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory model Poverty Prediction (n = 1000)"

@@

# Discussion {#sec-discussion}

@@

## Policy Implications
Broaden the discussion to include potential policy implications and recommendations based on your findings. How can policymakers use this information to address household poverty?

@@

### Lump sum nature of tax credit
In the SPM, taxes paid and tax credits in the year that are earned are include [@spm]d. But typically the credit would not appear in the family income until the next year. Also, tax credits are given as a lump sum, but poverty is a something that might happen anytime in the year. For example, a family may experience poverty because someone lost a job and cannot find another, so the family is in poverty. The tax credit may or may not provide relief depening on the timing in the year and the severity of expenses faced by the family throughout the year. So it is not very accurate that the SPM examine resources in a full calendar year compared with a threshold based on the calendar year. The family might have experienced 11 months of poverty before they got their tax credits in the 12th month.

## Future Steps
not done

@@

# Data Features

```{r}
#| label: tbl-feature
#| tbl-cap: "Preview of the raw 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false
#| warning: false

@@

## 90% Credibility Interval {#sec-credibility-interval}
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-modelresults1
#| fig-cap: "Credible intervals for predictors of positive poverty status"

>>>>

\newpage
# Introduction
Over the last few decades, the United States have implemented many policies, in an effort to address and mitigate poverty [@Kearney_Harris]. In 2019, the United States experienced the culmination of an economic expansion that followed the Great Recession. The poverty rate was estimated to be 10.5% in 2019 [@Semega_2022]. In order to reduce poverty further, it is important for us to analyze why some US household are in poverty in 2019. We analyze the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of official poverty measure critics, providing an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. We primarily look at two factors: mortgage status and income in each household. This allows us to look at the question whether potential financial factors may cause the poverty status for some US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various financial factors captured in the SPM dataset. We select 17 data features for analysis. After some data cleaning, we summarize 17 features into 2 variables: 1) mortgage state and 2) household annual total income after tax, plus government benefits and minus expenses.

@@

## Estimand
The estimand in this paper is the number of households who are in poverty in reality. However, it is difficult to measure the exact number of households who are in poverty since there are millions of people in the United States and not all of them will be assessed due to various difficulties. For example, due to regional issues, some people who live in rural areas are not included in the survey. Therefore, in this paper, we attempt to estimate the estimand using a logistic regression model which is fitted using sample dataset from the 2019 Revised Supplemental Poverty Measure Research Data from USCS.

@@

## Features
The original SPM 2019 dataset, which shows in @tbl-raw in Appendix @sec-data-details, contains 157959 data entries and many variables. Since it is difficult to observe such a large dataset, this report will only explore and analyze through several data features. We first chose these 17 variables: h_seq, spm_poor, spm_tenmortstatus, spm_totval, spm_snapsub, spm_caphousesub, spm_schlunch, spm_engval, spm_wicval, spm_fedtax, spm_eitc, spm_actc, spm_fica, spm_sttax, spm_childsuppd, spm_capwkccxpns, and spm_medxpns. Their explanation are in @tbl-feature. We can use those features (other than h_seq, spm_poor, spm_tenmortstatus) to calculate the income after tax, plus all the subsidies and benefits, minus all the expenses, just like how the SPM does it [@sgp2019]. This way the income is able to reflect the real financial situation on households more accurately.

@@

@@

The unit of measurement for numerical data are in US Dollars. For other data, they are categorical features with no units.

@@

First, a new column "income after tax" is created by summing up all incomes, credits, and benefits and subtracting all expenses. Then the dataset is cleaned by renaming of column headers, defining column classes, grouping the income variable into different income levels (here the assumed facts are that the income unit is in USD and 'k' stands for thousand, eg. 5k = 10,000), and replacing numerical values in the table with their corresponding descriptions from the data dictionary to improve the readability.

After cleaning, 62990 rows of data with 3 data features remain. @tbl-clean-data shows a preview of the cleaned dataset.

@@

```{r fig.pos="h"}
#| label: tbl-clean-data
#| tbl-cap: "Preview of the cleaned 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false

colnames(poverty2019) <- c("poverty status", "mortgage state", "income level")

@@

```{r fig.pos="h"}
#| label: tbl-summary
#| tbl-cap: "Statistics summary of the cleaned 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false

@@

```{r fig.pos="h"}
#| label: fig-mortgage
#| fig-cap: The distribution of poverty status by mortgage state
#| echo: false
#| warning: false
#| message: false

colnames(poverty2019) <- c("poverty_status", "mortgage_state", "income")
poverty2019 |>
  ggplot(aes(x = mortgage_state, fill = poverty_status)) +
  geom_bar(position = "dodge", width = 0.5) +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Mortgage state",
    y = "Number of households",
    fill = "Poverty Status"
  )
```

@@

```{r fig.pos="h"}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false

@@

@@

## Model Validation
For posterior predictive checks, in @fig-post_dist, the great fit of the posterior distribution from our logistic regression model with actual poverty data suggests accurate capture of poverty status patterns. This indicates the accuracy of our model's poverty status prediction to the 2019 SPM data. Also, @fig-post_prior compares the posterior to the prior, which shows some parameter changes such as "below 5k" and "5k-50k", as well as the intercept (people who are owners with mortgage and earn 100-150k). The discrepancy for the parameter suggest that our prior may not be very accurate in regards to these specific aspects.

More explanation can be found in Appendix @sec-model-details.
```{r fig.pos="h"}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory model Poverty Prediction (n = 1000)"

@@

\newpage
# Discussion {#sec-discussion}

@@

@@

### Lump sum nature of tax credit
In the SPM, taxes paid and tax credits in the year that are earned are include [@spm]d. But typically the credit would not appear in the family income until the next year. Also, tax credits are given as a lump sum, but poverty is a something that might happen anytime in the year. For example, a family may experience poverty because someone lost a job and cannot find another, so the family is in poverty. The tax credit may or may not provide relief depending on the timing in the year and the severity of expenses faced by the family throughout the year. So it is not very accurate that the SPM examine resources in a full calendar year compared with a threshold based on the calendar year. The family might have experienced 11 months of poverty before they got their tax credits in the 12th month.

## Future Steps
Future research should focus on gathering more inclusive and representative data by increasing the volume of data entries and mitigating bias in data collection. Beyond simply collecting data about household financial data, researchers should investigate more government benefits and expenses that are imposed on households. By investigating those benefits and expenses, researchers can arrive at conclusions if the current government poverty policies are effective for household in poverty. It is important to ensure that households in poverty have access to the necessary resources and support by the US government. This makes sure that the research not only analyzes the situation, but also contributes to actionable improvements for those affected.

@@

# Data {#sec-data-details}
## Raw data
@tbl-raw shows the preview of the raw 2019 US Supplemental Poverty Measure dataset.
```{r fig.pos="h"}
#| label: tbl-raw
#| tbl-cap: "Preview of the raw 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false

## Data feature
@tbl-feature shows the selected features of the 2019 US Supplemental Poverty Measure dataset.
```{r fig.pos="h"}
#| label: tbl-feature
#| tbl-cap: "Selected feature of the 2019 US Supplemental Poverty Measure dataset"
#| message: false
#| echo: false
#| warning: false

@@

## 90% Credibility Interval {#sec-credibility-interval}
@fig-modelresults1 and @fig-modelresults2 are 90% credibility interval plots.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-modelresults1
#| fig-cap: "Credible intervals for predictors of positive poverty status"
